Towards a standardized evaluation of imputation methodology

### 2022-09-19 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
update 2.3 and refs

Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) GV TEVEEL HAAKJES comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the measurement level, and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

@@

GV STREAMLINEN [TODO: maybe add definitions of missingness mechanisms back in???] A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly}. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}. Inducing a MAR mechanism presents the simulator with a choice, namely the type or functional form of the missingness model. Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model \citep{scho18}. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook such results due to the focus on a single functional form of the missingness-generating model.

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to emulate a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the true data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is generated randomly (see Figure 1). In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than imputation \citep[see e.g.][\S 2.7]{buur18}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{plots/plot_mmech.pdf}
  }      
  \caption{Three different realizations of missingness generation for MCAR and right-tailed MAR missingness mechanisms for different levels of correlation $\rho$ between the variable to ampute and the covariates that guide the probability to be missing. Note that the conditional probabilities of MCAR and right-tailed MAR ($rho=0$) are identical.}
    \label{Fig4}
\end{figure}

@@

In addition to varying the missingness mechanisms, each simulated mechanism should be combined with different missingness patterns. Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions becomes increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under several missingness proportions to amulate a realistic range in severity of the missingness problem. Depending on how the missingness mechanism interacts with the simulated data, higher missingness proportions may yield biased or invalid completed data inferences. The missingness proportions should therefore be considered carefully.

[TODO: paraphrase ``be clear which mechanisms are of interest, how they are generated, if they are suitably generated" as conclusion to this section.]
% \begin{table}[htb]
% \begin{center}
% \caption{Suggested univariate missingness proportions. [TODO: change this to 'more than one missingness proportion, depending on the context' and/or add references for the given proportions (e.g. about  that 30\% is a good range for MI).]}
% \label{table:prop}
% \begin{tabular}{lll}
% \hline
%       & Reasoning for suggested missingness proportion \\
% \hline  
% 10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
%       & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
%       & the true data inference. Performance of a missing data method should at least be \\
%       & acceptable for most missing data problems. \\ \noalign{\smallskip} % & \\
% 25\%  & This is a fair amount of missingness and will, depending on the observed data \\
%       & information, have a noticeable influence on the completed data inference. When \\
%       & compared to the condition with 10\% missingness, the inference obtained under \\
%       & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
%       & should increase), but estimates should still be properly covered and the statistical \\
%       & properties of the missing data method should be sound. In practice, at least to our \\
%       & at least to our experience in social sciences and official statistics, 25\% \\
%       & univariate missingness can easily be considered as a realistic missingness percentage. \\ \noalign{\smallskip} % & \\
% 50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
%       & severely. Depending on how the missingness mechanism interacts with the simulated \\
%       & data, some imputation techniques may yield estimates that are under-covered such that \\
%       & the completed data inference should not be deemed valid anymore. If a method yields \\
%       & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
%       & properties of the imputation methodology are sound. \\
% \hline
% \end{tabular}
% \end{center}
% \end{table}

>>>>

Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other potentially unfair comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the measurement level, and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

@@

Besides MCAR missingness, data dependent missingness mechanisms, such as missing at random (MAR) missingness are paramount in statistical inference investigations. A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data, generating the missingness through autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly}. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}.

Inducing a MAR mechanism presents the simulator with a choice, namely the type or functional form of the missingness model. Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model \citep{scho18}. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook such results due to the focus on a single functional form of the missingness-generating model.

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to mimic a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the true data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is generated randomly (see Figure 1). In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than imputation \citep[see e.g.][\S 2.7]{buur18}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

\begin{figure}[ht!]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{plots/plot_mmech.pdf}
  }      
  \caption{Three different realizations of missingness generation for MCAR and right-tailed MAR missingness mechanisms for different levels of correlation $\rho$ between the variable to ampute and the covariates that guide the probability to be missing. Note that the conditional probabilities of MCAR and right-tailed MAR ($rho=0$) are identical.}
    \label{Fig4}
\end{figure}

@@

In addition to varying the missingness mechanisms, each simulated mechanism should be combined with different missingness patterns. Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process \citep[Corollary 6.1A]{litt20}. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness --or if there is no data, for that matter-- the inference does not depend on the assumption.

Alternatively, the validity of assumptions becomes increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under several missingness proportions to amulate a realistic range in severity of the missingness problem. Depending on how the missingness mechanism interacts with the simulated data, higher missingness proportions may yield biased or invalid completed data inferences. The missingness proportions should therefore be considered carefully.

The above emphasizes the need for both thorough evaluation and thorough documentation of the intended and used missingness mechanisms. Simulators must be make clear which mechanisms are of interest. To be able to verify these mechanisms, simulators myst also be transparant about the process that generated the missingness. For example by sharing the code or detailing the process that generated the missing values. Alternatively, missingness generation devices, such as the \texttt{mice::ampute()} function can be used to generate valid, pattern-based missingness in any structured data set \citep{mice, ampute}. Finally, simulators must include simulation conditions and evaluation parameters that make it possible to identify the proper generation of missing values. This would prevent that errors in the missingness generation would go unnoticed.

% \begin{table}[htb]
% \begin{center}
% \caption{Suggested univariate missingness proportions. [TODO: change this to 'more than one missingness proportion, depending on the context' and/or add references for the given proportions (e.g. about  that 30\% is a good range for MI).]}
% \label{table:prop}
% \begin{tabular}{lll}
% \hline
%       & Reasoning for suggested missingness proportion \\
% \hline  
% 10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
%       & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
%       & the true data inference. Performance of a missing data method should at least be \\
%       & acceptable for most missing data problems. \\ \noalign{\smallskip} % & \\
% 25\%  & This is a fair amount of missingness and will, depending on the observed data \\
%       & information, have a noticeable influence on the completed data inference. When \\
%       & compared to the condition with 10\% missingness, the inference obtained under \\
%       & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
%       & should increase), but estimates should still be properly covered and the statistical \\
%       & properties of the missing data method should be sound. In practice, at least to our \\
%       & at least to our experience in social sciences and official statistics, 25\% \\
%       & univariate missingness can easily be considered as a realistic missingness percentage. \\ \noalign{\smallskip} % & \\
% 50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
%       & severely. Depending on how the missingness mechanism interacts with the simulated \\
%       & data, some imputation techniques may yield estimates that are under-covered such that \\
%       & the completed data inference should not be deemed valid anymore. If a method yields \\
%       & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
%       & properties of the imputation methodology are sound. \\
% \hline
% \end{tabular}
% \end{center}
% \end{table}
